---
title: "Clustering Analysis"
author: "Xiaoyan Wu"
email: xiaoyan.psych@gmail.com
date: "2026-02-20"
output: html_document
---

# Load packages and read dataset with cluster information
```{r}
rm(list=ls())
library(readr)
library(dplyr)
library(tidyr)
library(rstan)
library(purrr)
library(here)

cluster_path <- here("dataset", "subject_gender_cluster.csv")

cluster_df <- read_csv(
  cluster_path,
  col_types = cols(
    subid   = col_integer(),
    gender  = col_character(),   # "Male" / "Female"
    cluster = col_integer()
  )
) %>%
  rename(subject_id = subid,
         gender_label = gender)
cluster_df
```

# # Read behavioral data (trial-level)
```{r}
csv_path <- here("dataset", "trials_gender.csv")

dat <- read_csv(
  csv_path,
  col_types = cols(
    subject_id = col_integer(),
    gender     = col_character(),   # '男' / '女'
    trial      = col_integer(),
    action     = col_integer(),
    block      = col_integer(),
    violator   = col_double(),
    victim     = col_double(),
    cost       = col_double(),
    ratio      = col_double(),
    rt         = col_double()
  )
)

subjects <- sort(unique(dat$subject_id))
S <- length(subjects)
```

# Construct subject-level information table
```{r}
sub_df <- tibble(subject_id = subjects) %>%
  left_join(cluster_df, by = "subject_id")

# check missing cluster
table(is.na(sub_df$cluster))

print(sub_df)
```

# Extract posterior samples and compute subject-level posterior means (10 parameters)
```{r}

fit <- readRDS(
  here("Results", "fit_m08_SI_SCI_VCI_EC_RP_ID_lapse.rds")
)

post <- rstan::extract(fit)

param_names <- c(
  "gama","envy","guilt","lambda","kapa",
  "etak","etaa","oumiga","minp","maxp"
)

param_df <- map_dfc(
  param_names,
  ~ tibble(!! .x := colMeans(post[[.x]]))
) %>%
  mutate(subject_id = subjects) %>%
  left_join(sub_df, by = "subject_id")

head(param_df)
```

# Boys (2 clusters): Wilcoxon rank-sum test
```{r}

library(dplyr)
library(purrr)
library(tibble)

boys_df <- param_df %>%
  filter(gender_label == "Male")

wilcox_boys <- map_df(
  param_names,
  function(p) {
    x <- boys_df %>% filter(cluster == 1) %>% pull(p)
    y <- boys_df %>% filter(cluster == 2) %>% pull(p)

    w <- wilcox.test(x, y, exact = FALSE)

    # sample size
    n1 <- length(x)
    n2 <- length(y)
    N  <- n1 + n2

    # W statistic
    W <- as.numeric(w$statistic)

    # Normal approximation Z (Wilcoxon rank-sum test)
    mu_W  <- n1 * n2 / 2
    sd_W  <- sqrt(n1 * n2 * (n1 + n2 + 1) / 12)
    Z     <- (W - mu_W) / sd_W

    # Effect size r
    r <- Z / sqrt(N)

    tibble(
      parameter = p,
      test      = "Wilcoxon rank-sum",
      n_cluster1 = n1,
      n_cluster2 = n2,
      W          = round(W, 2),
      Z          = round(Z, 3),
      r          = round(r, 3),
      median_c1  = median(x, na.rm = TRUE),
      IQR_c1     = IQR(x, na.rm = TRUE),
      median_c2  = median(y, na.rm = TRUE),
      IQR_c2     = IQR(y, na.rm = TRUE),
      p_value    = w$p.value
    )
  }
) %>%
  mutate(
    p_holm  = p.adjust(p_value, method = "holm"),
    p_value = round(p_value, 6),
    p_holm  = round(p_holm, 6),
    median_c1 = round(median_c1, 3),
    IQR_c1    = round(IQR_c1, 3),
    median_c2 = round(median_c2, 3),
    IQR_c2    = round(IQR_c2, 3)
  )

cat("\n===== Boys (2 clusters): Wilcoxon + Holm (Full Report) =====\n")
print(wilcox_boys)

```


# Girls (three clusters): Kruskal–Wallis test with Holm correction
```{r}
library(dplyr)
library(purrr)
library(tibble)

girls_df <- param_df %>%
  filter(gender_label == "Female")

kw_girls <- map_df(
  param_names,
  function(p) {

    f <- as.formula(paste(p, "~ factor(cluster)"))
    k <- kruskal.test(f, data = girls_df)
    H  <- as.numeric(k$statistic)
    df <- as.numeric(k$parameter)   # df = k - 1 = 2
    p_val <- k$p.value

    # sample size
    n <- sum(!is.na(girls_df[[p]]))
    k_groups <- length(unique(girls_df$cluster))

    # Kruskal–Wallis eta-squared
    eta2_H <- (H - k_groups + 1) / (n - k_groups)

    tibble(
      parameter = p,
      test      = "Kruskal–Wallis",
      H         = round(H, 3),
      df        = df,
      eta2_H    = round(eta2_H, 3),
      p_value   = p_val
    )
  }
) %>%
  mutate(
    p_corr  = p.adjust(p_value, method = "bonferroni"),
    p_value = round(p_value, 6),
    p_corr  = round(p_corr, 6)
  )

cat("\n===== Girls (3 clusters): Kruskal–Wallis + Holm (main effects only) =====\n")
print(kw_girls)

```


# Girls: post hoc tests
```{r}
library(FSA)
library(dplyr)
library(purrr)

sig_params <- kw_girls %>%
  filter(p_corr < 1) %>%
  pull(parameter)

posthoc_girls <- map_df(
  sig_params,
  function(p) {

    f <- as.formula(paste(p, "~ factor(cluster)"))

    d <- dunnTest(
      f,
      data   = girls_df,
      method = "holm"   # Dunn's post hoc test with Holm correction
    )

    # Effective sample size for computing effect size r
    n <- sum(!is.na(girls_df[[p]]))

    d$res %>%
      transmute(
        Parameter  = p,
        Comparison = Comparison,
        Z          = round(Z, 3),
        r          = round(Z / sqrt(n), 3),
        p_corr     = round(P.adj, 6)   # p_holm_global
      )
  }
)

posthoc_girls <- posthoc_girls %>%
  mutate(
    p_bonf_global = round(p.adjust(p_corr, method = "bonferroni"), 6)
  )

print(posthoc_girls, n = Inf)
write.csv(posthoc_girls, "posthoc_girls_final.csv", row.names = FALSE)
```


# Standardize parameters across all subjects
```{r}
library(dplyr)

# -------------------------------------------------
# Apply min–max normalization to each parameter across all subjects
# -------------------------------------------------

param_df_scaled <- param_df

for (p in param_names) {
  min_p <- min(param_df[[p]], na.rm = TRUE)
  max_p <- max(param_df[[p]], na.rm = TRUE)

  param_df_scaled[[p]] <- (param_df[[p]] - min_p) / (max_p - min_p)
}

# -------------------------------------------------
# Compute parameter means by gender × cluster
# -------------------------------------------------

radar_df <- param_df_scaled %>%
  group_by(gender_label, cluster) %>%
  summarise(
    across(all_of(param_names), mean, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    gender_label = gender_label
  )
radar_df


boys_df  <- radar_df %>% filter(gender_label == "Male")
girls_df <- radar_df %>% filter(gender_label == "Female")


```


# Girls: radar plot
```{r}
library(fmsb)
library(scales)

# -----------------------------
# Parameter names (for radar plot display)
# -----------------------------
param_labels <- c(
  expression(gamma),
  expression(alpha),
  expression(beta),
  expression(lambda),
  expression(kappa),
  expression(eta[keep]),
  expression(eta[act]),
  expression(omega),
  expression(p[min]),
  expression(p[max])
)

# -----------------------------
# Girls: prepare data for radar chart
# -----------------------------
girls_mat <- girls_df %>%
  select(all_of(param_names)) %>%
  as.data.frame()

girls_radar <- rbind(
  rep(1, length(param_names)),  # max
  rep(0, length(param_names)),  # min
  girls_mat
)

colnames(girls_radar) <- param_names

# -----------------------------
# Girls: color palette
# -----------------------------
colors_girls <- c(
  "#B500B2",  # Cluster 1
  "#FF52A0",  # Cluster 2
  "#FFA47F"   # Cluster 3
)

cols_girls_line <- alpha(colors_girls, 0.75)
cols_girls_fill <- alpha(colors_girls, 0)

# -----------------------------
# Girls: radar chart
# -----------------------------
par(family = "Arial")
radarchart(
  girls_radar,
  axistype = 0,

  pcol  = cols_girls_line,
  plwd  = 3,
  plty  = 1,
  pfcol = cols_girls_fill,

  cglcol = "grey85",
  cglty  = 1,
  cglwd  = 0.8,

  axislabcol = "grey60",
  vlcex = 1,
  vlabels = param_labels,

  title = "Girls: Cluster-specific parameter profiles"
)

legend(
  "topright",
  legend = paste0("Cluster ", girls_df$cluster),
  col = colors_girls,
  lwd = 3,
  lty = 1,
  bty = "n"
)


# -----------------------------
# Export boys radar chart (PNG)
# -----------------------------

png(
  filename = "girls_radar.png",
  width = 1800,
  height = 1300,
  res = 300
)

par(family = "Arial")

radarchart(
  girls_radar,
  axistype = 0,

  pcol  = cols_girls_line,
  plwd  = 3,
  plty  = 1,
  pfcol = cols_girls_fill,

  cglcol = "grey85",
  cglty  = 1,
  cglwd  = 0.8,

  axislabcol = "grey60",
  vlcex = 1,
  vlabels = param_labels,

  title = "Girls: Cluster-specific parameter profiles"
)

legend(
  "topright",
  legend = paste0("Cluster ", sort(unique(girls_df$cluster))),
  col = colors_girls,
  lwd = 3,
  lty = 1,
  bty = "n"
)

dev.off()
```


# Girls: boxplot
```{r}
#=================================
# Data preparation
#=================================
girls_mat <- param_df %>% 
  filter(gender_label == "Female")

library(dplyr)
library(tidyr)
library(ggplot2)

#---------------------------------
# 1. Parameters to plot
#---------------------------------
params_plot <- c("gama", "envy", "kapa","etak", "oumiga", "minp")

#---------------------------------
# 2. Long format
#---------------------------------
girls_long <- girls_mat %>%
  select(cluster, all_of(params_plot)) %>%
  pivot_longer(
    cols = all_of(params_plot),
    names_to = "parameter",
    values_to = "value"
  ) %>%
  mutate(
    cluster = factor(cluster)
  )

#---------------------------------
# 3. Colors
#---------------------------------
colors_girls <- c(
  "1" = "#B500B2",
  "2" = "#FF52A0",
  "3" = "#FFA47F"
)

#---------------------------------
# 4. Parameter labels
#---------------------------------
param_labels <- c(
  gama   = "gamma",
  envy   = "alpha",
  lambda = "lambda",
  kapa   = "kappa",
  etak = "etak",
  oumiga = "omega",
  minp   = "p[min]"
)

#---------------------------------
# 5. Plot
#---------------------------------
p_girls <- ggplot(
  girls_long,
  aes(x = cluster, y = value, color = cluster)
) +
  geom_boxplot(
    width = 0.6,
    outlier.shape = NA,
    linewidth = 1
  ) +
  geom_jitter(
    width = 0.15,
    size = 2,
    alpha = 0.4
  ) +
  facet_wrap(
    ~ parameter,
    scales = "free_y",
    ncol = 6,
    labeller = as_labeller(param_labels, label_parsed)
  ) +
  scale_color_manual(values = colors_girls) +
  labs(
    x = "Cluster",
    y = "Parameter value",
    title = "Girls: Cluster differences in model parameters",
    subtitle = "Boxplots show median and IQR; points represent individual participants"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "none",
    strip.text = element_text(size = 14, face = "bold"),
    plot.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(size = 13),

    # ⭐ 与 boys 图完全一致的关键设置
    panel.grid = element_blank(),                    # 去掉灰色 grid
    axis.line = element_line(color = "black", linewidth = 0.8),
    axis.ticks = element_line(color = "black")
  )

#---------------------------------
# 6. Display
#---------------------------------
print(p_girls)

ggsave(
  "girls_cluster_parameters.png",
  p_girls,
  width = 14,
  height = 4,
  dpi = 300
)
```

# Boys: radar chart
```{r}
# -----------------------------
# Boys: prepare data for radar chart
# -----------------------------

param_labels <- c(
  gama   = expression(gamma),
  envy   = expression(alpha),
  guilt  = expression(beta),
  lambda = expression(lambda),
  kapa   = expression(kappa),
  etak   = expression(eta[keep]),
  etaa   = expression(eta[action]),
  oumiga = expression(omega),
  minp   = expression(p[min]),
  maxp   = expression(p[max])
)
boys_mat <- boys_df %>%
  select(all_of(param_names)) %>%
  as.data.frame()

boys_radar <- rbind(
  rep(1, length(param_names)),  # max
  rep(0, length(param_names)),  # min
  boys_mat
)

colnames(boys_radar) <- param_names


# -----------------------------
# Boys: color palette
# -----------------------------
colors_boys <- c(
  "#685AFF",   # Cluster 1
  "#9CCFFF"  # Cluster 2
)

cols_boys_line <- alpha(colors_boys, 0.75)
cols_boys_fill <- alpha(colors_boys, 0)

# -----------------------------
# Boys: radar chart
# -----------------------------
par(family = "Arial")
radarchart(
  boys_radar,
  axistype = 0,

  pcol  = cols_boys_line,
  plwd  = 3,
  plty  = 1,
  pfcol = cols_boys_fill,

  cglcol = "grey85",
  cglty  = 1,
  cglwd  = 0.8,

  axislabcol = "grey60",
  vlcex = 1,
  vlabels = param_labels,

  title = "Boys: Cluster-specific parameter profiles"
)

legend(
  "topright",
  legend = paste0("Cluster ", boys_df$cluster),
  col = colors_boys,
  lwd = 3,
  lty = 1,
  bty = "n"
)


# -----------------------------
# Export boys radar chart (PNG)
# -----------------------------
png(
  filename = "boys_radar.png",
  width = 1800,
  height = 1300,
  res = 300
)

par(family = "Arial")

radarchart(
  boys_radar,
  axistype = 0,

  pcol  = cols_boys_line,
  plwd  = 3,
  plty  = 1,
  pfcol = cols_boys_fill,

  cglcol = "grey85",
  cglty  = 1,
  cglwd  = 0.8,

  axislabcol = "grey60",
  vlcex = 1,
  vlabels = param_labels,

  title = "Boys: Cluster-specific parameter profiles"
)

legend(
  "topright",
  legend = paste0("Cluster ", sort(unique(boys_df$cluster))),
  col = colors_boys,
  lwd = 3,
  lty = 1,
  bty = "n"
)

dev.off()
```


# Boys: boxplot
```{r}
#=================================
# Data preparation
#=================================
boys_mat <- param_df %>% 
  filter(gender_label == "Male")

library(dplyr)
library(tidyr)
library(ggplot2)

#---------------------------------
# 1. Parameters to plot
#---------------------------------
params_plot_boys <- c("gama", "envy", "kapa", "minp", "maxp")

#---------------------------------
# 2. Long format
#---------------------------------
boys_long <- boys_mat %>%
  select(cluster, all_of(params_plot_boys)) %>%
  pivot_longer(
    cols = all_of(params_plot_boys),
    names_to = "parameter",
    values_to = "value"
  ) %>%
  mutate(
    cluster = factor(cluster)
  )

#---------------------------------
# 3. Colors
#---------------------------------
colors_boys <- c(
  "1" = "#685AFF",
  "2" = "#9CCFFF"
)

#---------------------------------
# 4. Parameter labels
#---------------------------------
param_labels_boys <- c(
  gama = "gamma",
  envy = "alpha",
  kapa = "kappa",
  minp = "p[min]",
  maxp = "p[max]"
)

#---------------------------------
# 5. Plot
#---------------------------------
p_boys <- ggplot(
  boys_long,
  aes(x = cluster, y = value, color = cluster)
) +
  geom_boxplot(
    width = 0.6,
    outlier.shape = NA,
    linewidth = 1
  ) +
  geom_jitter(
    width = 0.15,
    size = 2,
    alpha = 0.4
  ) +
  facet_wrap(
    ~ parameter,
    scales = "free_y",
    ncol = 5,
    labeller = as_labeller(param_labels_boys, label_parsed)
  ) +
  scale_color_manual(values = colors_boys) +
  labs(
    x = "Cluster",
    y = "Parameter value",
    title = "Boys: Cluster differences in model parameters",
    subtitle = "Boxplots show median and IQR; points represent individual participants"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "none",
    strip.text = element_text(size = 14, face = "bold"),
    plot.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(size = 13),

    # ⭐ 关键三行
    panel.grid = element_blank(),                    # 去掉灰色 grid
    axis.line = element_line(color = "black", linewidth = 0.8),  # 加回 x/y 轴线
    axis.ticks = element_line(color = "black")       #（可选）轴刻度更清晰
  )

#---------------------------------
# 6. Display & save
#---------------------------------
print(p_boys)

ggsave(
  "boys_cluster_parameters.png",
  p_boys,
  width = 14,
  height = 4,
  dpi = 300
)
```