---
title: "Model Compasion Analysis"
author: "Xiaoyan Wu"
email: xiaoyan.psych@gmail.com
date: "2026-02-20"
output: html_document
---

# 导入所有模型
```{r}
rm(list=ls())
require("knitr") 
suppressMessages(library('rstan'))
suppressMessages(library('loo'))
suppressMessages(library('parallel'))
suppressMessages(library('hBayesDM'))
suppressMessages(library("reshape2"))
library(StanHeaders)
library(ggplot2)
library(rstan)
library(loo)
library(reshape2)
library(hBayesDM)
library(readxl)
library(readr)
library(dplyr)
library(tidyr)
library(here)
rstan_options(auto_write = TRUE)
options(mc.cores = parallel::detectCores())
library(purrr)

csv_path <- here("dataset", "trials_gender.csv")

dat <- read_csv(csv_path,
                col_types = cols(
                  subject_id = col_integer(),
                  gender     = col_character(),
                  trial      = col_integer(),
                  action     = col_integer(),
                  block      = col_integer(),
                  violator   = col_double(),
                  victim     = col_double(),
                  cost       = col_double(),
                  ratio      = col_double(),
                  rt         = col_double()
                ))

dat <- dat %>%
  mutate(gender_num = ifelse(gender == "女", 1L, 2L)) %>%
  arrange(subject_id, trial)

subjects <- unique(dat$subject_id)
S <- length(subjects)

group <- integer(S)

for (i in seq_along(subjects)) {
  sid <- subjects[i]
  df  <- dat %>% filter(subject_id == sid)
  group[i] <- df$gender_num[1]
}

fit_dir <- here("Results")

fit_files <- list.files(
  fit_dir,
  pattern = "^fit_.*\\.rds$",
  full.names = TRUE
)

fits <- map(fit_files, readRDS)
names(fits) <- gsub("^fit_|\\.rds$", "", basename(fit_files))

```


# Model comparision
```{r}
#--------------------------------------------------
# Extract subject-level log-likelihood
#--------------------------------------------------

loglik_subj_list <- map(
  fits,
  ~ {
    ll <- rstan::extract(
      .x,
      pars = "log_lik_subj",
      permuted = FALSE
    )
    # ll: iterations × chains × S
    apply(ll, c(1, 3), sum)
  }
)

#--------------------------------------------------
# 3. PSIS-LOO（subject-level）
#--------------------------------------------------

loo_list <- map(loglik_subj_list, loo::loo)
names(loo_list) <- names(fits)

#--------------------------------------------------
# 4. 模型比较
#--------------------------------------------------

loo_compare_res <- loo::loo_compare(loo_list)
print(loo_compare_res)

```

# Model comparison across multiple metrics
```{r}
## PSIS-LOO

# Extract ELPD and LOOIC
loo_summary <- map_df(
  loo_list,
  ~ tibble(
    elpd_loo = .x$estimates["elpd_loo", "Estimate"],
    se_elpd  = .x$estimates["elpd_loo", "SE"],
    looic    = .x$estimates["looic", "Estimate"],
    se_looic = .x$estimates["looic", "SE"]
  ),
  .id = "model"
)

loo_summary


## WAIC
waic_list <- map(loglik_subj_list, loo::waic)

waic_summary <- map_df(
  waic_list,
  ~ tibble(
    waic   = .x$estimates["waic", "Estimate"],
    se_waic = .x$estimates["waic", "SE"]
  ),
  .id = "model"
)

waic_summary


## DIC
dic_list <- map(
  loglik_subj_list,
  ~ {
    dev <- -2 * .x
    D_bar <- mean(rowSums(dev))
    D_hat <- sum(colMeans(dev))
    DIC <- 2 * D_bar - D_hat
    DIC
  }
)

dic_df <- tibble(
  model = names(dic_list),
  DIC   = unlist(dic_list)
)

dic_df
```


# Perform model comparison separately by gender
```{r}
library(rstan)
library(loo)
library(purrr)
library(dplyr)
library(tibble)

#--------------------------------------------------
# Extract subject-level log-likelihood
#    (iterations × S)
#--------------------------------------------------

loglik_subj_list <- map(
  fits,
  ~ {
    ll <- rstan::extract(
      .x,
      pars = "log_lik_subj",
      permuted = FALSE
    )
    # ll: iterations × chains × S
    apply(ll, c(1, 3), sum)
  }
)

#--------------------------------------------------
# Gender index
#--------------------------------------------------

female_idx <- which(group == 1)
male_idx   <- which(group == 2)

#--------------------------------------------------
# Compute PSIS-LOO separately by gender
#--------------------------------------------------

loo_female <- map(
  loglik_subj_list,
  ~ loo::loo(.x[, female_idx, drop = FALSE])
)

loo_male <- map(
  loglik_subj_list,
  ~ loo::loo(.x[, male_idx, drop = FALSE])
)

#--------------------------------------------------
# Compute WAIC separately by gender
#--------------------------------------------------

waic_female <- map(
  loglik_subj_list,
  ~ loo::waic(.x[, female_idx, drop = FALSE])
)

waic_male <- map(
  loglik_subj_list,
  ~ loo::waic(.x[, male_idx, drop = FALSE])
)

#--------------------------------------------------
# 5. Extract LOO / WAIC metrics
#--------------------------------------------------

extract_loo_waic <- function(loo_list, waic_list, gender_label) {
  map_df(
    names(loo_list),
    ~ tibble(
      model   = .x,
      elpd_loo = loo_list[[.x]]$estimates["elpd_loo", "Estimate"],
      se_elpd  = loo_list[[.x]]$estimates["elpd_loo", "SE"],
      looic    = loo_list[[.x]]$estimates["looic", "Estimate"],
      se_looic = loo_list[[.x]]$estimates["looic", "SE"],
      waic     = waic_list[[.x]]$estimates["waic", "Estimate"],
      se_waic  = waic_list[[.x]]$estimates["waic", "SE"],
      gender   = gender_label
    )
  )
}

loo_waic_female_df <- extract_loo_waic(loo_female, waic_female, "Female")
loo_waic_male_df   <- extract_loo_waic(loo_male,   waic_male,   "Male")

loo_waic_df <- bind_rows(loo_waic_female_df, loo_waic_male_df)

#--------------------------------------------------
# 6. Compute DIC separately by gender
#--------------------------------------------------

compute_dic <- function(loglik_mat) {
  dev <- -2 * loglik_mat
  D_bar <- mean(rowSums(dev))
  D_hat <- sum(colMeans(dev))
  2 * D_bar - D_hat
}

dic_female <- map_dbl(
  loglik_subj_list,
  ~ compute_dic(.x[, female_idx, drop = FALSE])
)

dic_male <- map_dbl(
  loglik_subj_list,
  ~ compute_dic(.x[, male_idx, drop = FALSE])
)

dic_df <- bind_rows(
  tibble(model = names(dic_female),
         DIC = dic_female,
         gender = "Female"),
  tibble(model = names(dic_male),
         DIC = dic_male,
         gender = "Male")
)

#--------------------------------------------------
# 7. Combine all metrics
#--------------------------------------------------

model_comparison_gender <- loo_waic_df %>%
  left_join(dic_df, by = c("model", "gender")) %>%
  arrange(gender, looic)
```

```{r}
#--------------------------------------------------
# 8. Print results separately by gender
#--------------------------------------------------

model_comp_female <- model_comparison_gender %>%
  filter(gender == "Female") %>%
  arrange(looic)

model_comp_male <- model_comparison_gender %>%
  filter(gender == "Male") %>%
  arrange(looic)

cat("\n===== Female participants =====\n")
print(model_comp_female)

```

```{r}
cat("\n===== Male participants =====\n")
print(model_comp_male)
```

# Figures reported in the manuscript
# All subjects
```{r}
library(dplyr)
library(ggplot2)
library(tibble)

loo_df <- as.data.frame(loo_compare_res) %>%
  rownames_to_column("model") %>%
  mutate(
    model = factor(model)  # 最优模型在最上
  )
p_main <- ggplot(
  loo_df,
  aes(x = elpd_diff, y = model)
) +
  geom_vline(
    xintercept = 0,
    linetype = "dashed",
    color = "grey50",
    linewidth = 0.7
  ) +
  geom_point(size = 3) +
  geom_errorbarh(
    aes(
      xmin = elpd_diff - se_diff,
      xmax = elpd_diff + se_diff
    ),
    height = 0.25,
    linewidth = 0.8
  ) +
  labs(
    x = expression(Delta * "ELPD (relative to best model)"),
    y = NULL,
    title = "Model comparison using PSIS-LOO",
    subtitle = "Points indicate ΔELPD; error bars show ±1 SE"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    panel.grid.major.y = element_blank(),
    plot.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(size = 13),
    axis.text.y = element_text(size = 12)
  )

print(p_main)
```

# Plot figures separately by gender: Girls
```{r}
model_order <- model_comp_female %>%
  arrange(looic) %>%        # 或 arrange(elpd_loo)
  pull(model)

#--------------------------------------------------
# Model order
#--------------------------------------------------
model_order <- c(
  "m01_baseline",
  "m02_SI",
  "m03_SI_SCI",
  "m04_SI_SCI_VCI",
  "m05_SI_SCI_VCI_EC",
  "m06_SI_SCI_VCI_EC_RP",
  "m07_SI_SCI_VCI_EC_RP_ID",
  "m08_SI_SCI_VCI_EC_RP_ID_lapse",
  "m09_Heuristic",
  "m10_Heuristic_lapse"
)

best_elpd_female <- max(model_comp_female$elpd_loo)

#--------------------------------------------------
# plot（girls）
#--------------------------------------------------
p_female <- model_comp_female %>%
  mutate(
    model = factor(model, levels = model_order),
    elpd_diff = elpd_loo - best_elpd_female
  ) %>%
  ggplot(aes(x = elpd_diff, y = model)) +

  # reference line
  geom_vline(
    xintercept = 0,
    linetype = "dashed",
    color = "grey50",
    linewidth = 0.7
  ) +

  # points
  geom_point(
    size = 3,
    color = rgb(231, 113, 160, maxColorValue = 255)
  ) +

  # error bars (±1 SE)
  geom_errorbarh(
    data = ~ subset(.x, elpd_diff != 0),
    aes(
      xmin = elpd_diff - se_elpd,
      xmax = elpd_diff + se_elpd
    ),
    height = 0.25,
    linewidth = 0.8,
    color = rgb(231, 113, 160, maxColorValue = 255)
  ) +

  labs(
    x = expression(Delta * ELPD[LOO]),
    y = NULL,
    title = "Model comparison (Female participants)",
    subtitle = "PSIS-LOO; points indicate ΔELPD, error bars show ±1 SE"
  ) +

  theme_minimal(base_size = 14) +
  theme(
    panel.grid.major.y = element_blank(),
    plot.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(size = 13),
    axis.text.y = element_text(size = 12)
  )

print(p_female)

```


# Plot figures separately by gender: Boys
```{r}
library(dplyr)
library(ggplot2)

#--------------------------------------------------
# plot
#--------------------------------------------------
p_male <- model_comp_male %>%
  mutate(
    model = factor(model, levels = model_order),
    elpd_diff = elpd_loo - max(elpd_loo)
  ) %>%
  ggplot(aes(x = elpd_diff, y = model)) +
  
  geom_vline(
    xintercept = 0,
    linetype = "dashed",
    color = "grey50",
    linewidth = 0.7
  ) +
  
  geom_point(
    size = 3,
    color = rgb(93, 160, 198, maxColorValue = 255)   # Boys
  ) +
  
  geom_errorbarh(
    data = ~ subset(.x, elpd_diff != 0),
    aes(
      xmin = elpd_diff - se_elpd,
      xmax = elpd_diff + se_elpd
    ),
    height = 0.25,
    linewidth = 0.8,
    color = rgb(93, 160, 198, maxColorValue = 255)
  ) +
  
  labs(
    x = expression(Delta * "ELPD"[LOO]),
    y = NULL,
    title = "Model comparison (Male participants)",
    subtitle = "PSIS-LOO; error bars indicate ±1 SE"
  ) +
  
  theme_minimal(base_size = 14) +
  theme(
    panel.grid.major.y = element_blank(),
    plot.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(size = 13),
    axis.text.y = element_text(size = 12)
  )

print(p_male)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
